What We Eval in the Shadows

### 2021-08-13 paper/main.tex
Fix grammar issues and turn on review mode

\documentclass[screen,acmsmall]{acmart}%\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

@@

\paragraph{Controlled Evaluation}
\Eval can be used to control when and if expressions are evaluated. For
example, \k{assertthat} provides users with the ability to make assertions and,
if they fail, produce legible error messages. The example here accepts a vararg
with a list of expressions. It extracts the associated code in the same way
as the extracting varags example and evaluates each of the expression
sequentially in the caller's environment. If an expression does not hold, an
error is generated.

@@

\paragraph{Shadowing}
%%% TODO ?????????????//// <- left by Jan
Many packages use \eval to implement variations of the \k{base} R functions.
For example, the \k{igraph} package provides the \k{do_call} function shown
below. This is a variant of the builtin \k{do.call} function with the
additional support for specifying the environment where the call should happen.
It constructs a call from a name and a list of arguments, and pass it to \eval.

@@

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is expended
via \eval and \k{match.call}. The example uses \k{match.call} to access the
expression with which the function is called. The next three lines construct a
call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
\k{'data'}, and \k{'weights'}. Then, the call expression is evaluated in the
parent environment. The use of \k{match.call} avoids repeated use of
\k{substitute} for extracting the argument expressions. It is extensible as new
parameter names can be easily added. This example illustrates one of the most
recurring uses of \eval by packages that do statistical modeling, \ie the
\c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.

@@

This combination of \k{match.call} and \eval for passing most of the current
call's argument to another function is also used to created wrappers. For
example, the \k{write.csv} uses it to forward most of the parameters to the
more general \k{write.table} function, only setting the CSV-related
parameters:

@@

The most probable reason for this is to ease maintenance of such
wrappers.\footnote{Private communication with a developer of R.} Wrapping
functions in R is somewhat harder than in other languages because of default
arguments, varargs and missing arguments.

@@

\Eval allows functions to resolve symbols in different environments than the
current one. For example, the \k{data.table} redefines the subsetting
operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
  \texttt{j} is used to calculate a subset of rows from \texttt{i} grouped by \texttt{k}.}
for data frames to enable compact queries. The following  
computes the average departure delays on 2014 Delta flights for each
origin airport.

@@

At its core, the \k{[} operator uses \k{substitute} to capture its operand
expressions, transforms them, and evaluates them using \eval in the given data
frame.\footnote{We omit the code as the implementation spans over 1700 lines,\\
  \indent \cf
\url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}} The
symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up
in the current environment, but instead in the \k{flights} data frame.

The non-standard scoping is an important use case in R. It allows ergonomic
functions for interactive data exploration. The ergonomics come from the
reduced typing and syntactic noise at the cost of some ambiguity. Since R is
used primarily interactively, this trade-off is justifiable.

\paragraph{Domain-Specific Language}
A domain-specific language leverages R's grammar but redefines it semantics in
some suitable way. The usage example illustrates string interpolation. The
\k{glue} function from the \k{glue} package extracts snippets of code enclosed
between braces and evaluates them using \eval, and finally splices their
results to construct the result.

@@

The \k{glue} package contains a variant that leverages the non-standard scoping
allowing to resolve interpolation in an alternative environment. For example,
the following snippet pipes the subsetted data frame into the \k{glue_data}
function that uses it to resolve the symbols in string interpolation.

@@

\paragraph{Obfuscation}
A few packages use \eval to bypass the restrictions imposed by the \k{R CMD
  CHECK} tool. This tool enforces some well-formedness rules on a submitted
package before accepting it for inclusion in CRAN. Some static checks ensure
that the package code does not use certain restricted functions. Package authors
use \eval to obfuscate their code. The example mutates the variable \k{s} in
the package environment. However, that is locked by default and
\k{unlockBinding}, is a restricted function. To work around this, the call to
unlock is done through \eval.

@@

In R, \eval is used chiefly for meta-programming and access remote environments.
Unlike, JavaScript which had a small set of well-defined patterns that could be
rewritten without \eval, our results suggest that \eval is an integral part of
programming in R.

The cases where \eval appears to be overkill include simple expression shapes
such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
environment can be performed by the builtin \k{get} function, a similar case
can be made for assignment and the \k{assign} function. There is a dedicated
\k{do.call} function to perform function calls. However, the unnecessary uses
of \eval are not easy to find and often can only be uncovered by manual
analysis. Dynamic analysis is limited due to coverage issues, and static
analysis of R remains an open problem.

@@

For example in Listing~\ref{lst:unnecessary}, the function uses \eval to lookup
a target function name from a constructed string. Then it constructs a calls it
with the arguments that are passed to the current function. \Eval is overkill
in this case: a semantically equivalent function body can use \k{get} to perform
the lookup and then call the resulting function directly.

@@

Another example, from \k{coin::copyslots}, constructs an expression that access
fields of an object and copy them.

@@

We have not yet touched on alternative implementations of \eval. The tidyverse
libraries\footnote{An opinionated collection of R packages designed for data
science. It consists of some of the most popular packages.} include their own
implementation calls \k{rlang::eval\_tidy}~\cite{tidyverse}. This departs from
\eval in two ways. First, it supports evaluation of quosures, custom objects
used by for metaprogramming that bundle expressions with an environment.
\k{eval\_tidy} accepts a data mask argument, a set of bindings such as a
dataframe or list, which takes precedence over the environment. Effects such as
assignments happen in the data mask. Expressions such as \k{return()} do not
work since the data mask does not correspond to a frame on the call stack.

>>>>

\documentclass[review,screen,acmsmall]{acmart}%\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

@@

\paragraph{Controlled Evaluation}
\Eval can be used to control when and if expressions are evaluated. For
example, \k{assertthat} provides users with the ability to make assertions and,
if they fail, produce legible error messages. The example here accepts a vararg
with a list of expressions. It extracts the associated code in the same way
as the extracting varags example and evaluates the expressions
sequentially in the caller's environment. If an expression does not hold, an
error is generated.

@@

\paragraph{Shadowing}
%%% TODO ?????????????//// <- left by Jan
Some packages use \eval to implement variations of the \k{base} R functions.
For example, the \k{igraph} package provides the \k{do_call} function shown
below. This is a variant of the builtin \k{do.call} function with different
semantics for resolving the function name.
%with the additional support for specifying the environment where the call
%should happen. -- wrong, do.call also allows specifying an envir
It constructs a call from a name and a list of arguments, and passes it to \eval.

@@

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is expended
via \eval and \k{match.call}. The example uses \k{match.call} to access the
expression with which the function is called. The next three lines construct a
call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
\k{'data'}, and \k{'weights'}. Then, the call expression is evaluated in the
parent environment. The use of \k{match.call} avoids repeated use of
\k{substitute} for extracting the argument expressions. It is extensible as new
parameter names can be easily added. This example illustrates one of the most
recurring uses of \eval by packages that do statistical modeling, and relates to
the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.

@@

This combination of \k{match.call} and \eval for passing most of the current
call's argument to another function is also used to create wrappers. For
example, the \k{write.csv} uses it to forward most of the parameters to the
more general \k{write.table} function, only setting the CSV-related
parameters:

@@

The pattern eases maintenance of wrappers by avoiding the need to explicitly
list all the parameter names.\footnote{Private communication with a developer of
  R.} Wrapping functions in R is somewhat harder than in other languages because
of default arguments, varargs and missing arguments.

@@

\Eval allows functions to resolve symbols in different environments than the
current one. For example, the \k{data.table} package redefines the subsetting
operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
  \texttt{j} is used to calculate a subset of rows from \texttt{i} grouped by \texttt{k}.}
for data frames to enable compact queries. The following  
computes the average departure delays on 2014 Delta flights for each
origin airport.

@@

At its core, the \k{[} operator uses \k{substitute} to capture its operand
expressions, transforms them, and evaluates them using \eval in the given data
frame.\footnote{We omit the code as the implementation spans over 1700 lines,\\
  \indent \cf
\url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}} The
symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up
in the current environment, but in the \k{flights} data frame.

Non-standard scoping is an important use case of \eval in R. It enables the
design of functions which require less typing. Since R is used primarily
interactively, this trade-off is justifiable.

\paragraph{Domain-Specific Language}
A domain-specific language leverages R's grammar but redefines its semantics in
a suitable way. The example below illustrates string interpolation. The
\k{glue} function from the \k{glue} package extracts snippets of code enclosed
between braces and evaluates them using \eval, and finally splices their
results to construct the result.

@@

The \k{glue} package contains a variant of \k{glue} function that leverages
non-standard scoping for interpolation in a custom environment. For example, the
following snippet pipes the subsetted data frame into the \k{glue_data} function
that uses it to resolve the symbols for string interpolation.

@@

\paragraph{Obfuscation}
A few packages use \eval to bypass the restrictions imposed by the \k{R CMD
  CHECK} tool. This tool enforces some well-formedness rules on a submitted
package before accepting it for inclusion in CRAN. Some static checks ensure
that the package code does not use certain restricted functions. Package authors
use \eval to obfuscate their code to get around this limitation. The example
below mutates the variable \k{s} in the package environment. However, that is
locked by default and \k{unlockBinding} is a restricted function. To work around
this, the call to unlock is done through \eval.

@@

In R, \eval is used chiefly for meta-programming and accessing remote
environments. Unlike JavaScript which had a small set of well-defined patterns
that could be rewritten without \eval, our results suggest that \eval is an
integral part of programming in R.

The cases where \eval appears to be an overkill include simple expression shapes
such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
environment can be performed by the builtin \k{get} function, a similar case can
be made for assignment and the \k{assign} function. There is a dedicated
\k{do.call} function to perform function calls. However, the unnecessary uses of
\eval are not easy to find and often can only be uncovered by manual analysis.
Dynamic analysis is limited due to coverage issues, and static analysis of R
remains an open problem.

@@

For the example in Listing~\ref{lst:unnecessary}, the function uses \eval to
lookup a target function name from a constructed string. Then it constructs a
call with the arguments that are passed to the current function. \Eval is
overkill in this case: a semantically equivalent function body can use \k{get}
to perform the lookup and then call the resulting function directly.

@@

Another example, from \k{coin::copyslots}, constructs an expression that accesses
fields of an object and copies them.

@@

We have not yet touched on alternative implementations of \eval. The tidyverse
libraries\footnote{An opinionated collection of R packages designed for data
science. It consists of some of the most popular packages.} include their own
implementation calls \k{rlang::eval\_tidy}~\cite{tidyverse}. This departs from
\eval in two ways. First, it supports evaluation of quosures, custom objects
used for metaprogramming that bundle expressions with an environment.
\k{eval\_tidy} accepts a data mask argument, a set of bindings such as a
dataframe or list, which takes precedence over the environment. Effects such as
assignments happen in the data mask. Expressions such as \k{return()} do not
work since the data mask does not correspond to a frame on the call stack.
